---
title: "607 Assignment 5B Approach"
author: "Ony"
date: "2026-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 5B Approach

#ELO Chess Player Performance Evaluations

1.  Import the final dataframe from Project 1

2.  Use the Player pre-rating and opponent pre-rating data to calculate their expected score (potential formula: '1/(1+10\^((RB - RA)/400))').

3.  Compare this score to the actual score in a table, sort the players in descending order of discrepancy in expected and actual results, and isolate the top 5 and bottom 5.

## Execution

First, load the R.Data work-space created in project1. (The link below uses he local path as github couldn't provide a raw link for the RData file. Github link for download is: <https://github.com/ifi-ony/607_Assignment_5/blob/ca05c60f8f69414baa62fb165506534cd2830f28/project1workspace.RData>)

```{r}
load("C:/Users/tesla/Desktop/CUNY DATA SCIENCE DEGREE/607_Data_Aquisition/project_1/project1workspace.RData")
```

The code chunk below is copied from the Project 1 code, and it creates a list of vectors containing the pre-rating of every opponent faced by that player.

```{r create-opponent-list}
opps_no_na <- lapply(num_opplist, Filter, f = Negate(is.na))
opp_scores <- function(x) {
  num_prerate[unlist(opps_no_na[x])]
}

no_avg_opp_score <- lapply(player_num, opp_scores)

```

Next we use nested for loops to calculate every player's expected score against each of their opponents from the previous list. The resulting dataframe contains every player, their pre-rating, the ratings for each opponent they faced, their expected scores, their actual scores and the difference between the two.

```{r create-large-dataframe}


expected_scores <- vector(mode = "list", length = length(no_avg_opp_score))

overall_expected_score <- list()

for (i in 1:length(no_avg_opp_score)) {
  for (j in 1:length(no_avg_opp_score[[i]])) {
    prob <- (1/(1 + 10^((no_avg_opp_score[[i]][j] - full_frame$num_prerate[i])/400) ))
    
      expected_scores[[i]][j] = round(prob, digits = 2)
      
      overall_expected_score[i] = sum(expected_scores[[i]])
    
  }
}

opp_frame <- full_frame %>%
  mutate(opp_rating_list = no_avg_opp_score,
         expected_scores = expected_scores,
         overall_expected_score = overall_expected_score,
         overall_actual_score = points,
         real_minus_expected = overall_actual_score - as.numeric(overall_expected_score))
 


```

Finally, we sort the dataframe in descending order of score differential and separate it into two tables, one for the top 5 and the other for the bottom 5.

```{r }
cleaner_frame <- opp_frame %>%
  select(player_num, player_name, real_minus_expected) %>%
  arrange(desc(real_minus_expected))

overachievers <- head(cleaner_frame, 5) %>%
  gt() %>%
  tab_header("Overachievers List") %>%
  cols_label(player_name ~ "Player Name", player_num ~ "Player Number", real_minus_expected ~ "Score Delta")

better_luck_next_time <- tail(cleaner_frame, 5) %>%
  gt() %>%
  tab_header("Better Luck Next Time") %>%
  cols_label(player_name ~ "Player Name", player_num ~ "Player Number", real_minus_expected ~ "Score Delta")

overachievers
better_luck_next_time

```
